---
title:  'Activity: Popular Names'
author: "Dave Friedman"
date:   "05 June 2023"
output: html_notebook
---

# Grading

Assignment is worth a total of 10 points

* [1 point] Turn in BOTH RNotebook (.nb.html) and the RMarkdown (.Rmd) file to Canvas
* [1 point] Changes pushed to GitHub
* [1 point] Step 1: show how you used summary functions to explore the raw data set
* [1 point] Step 3 and 4: attempt to answer all questions about the data
* [1 point] Step 5: explanation of data wrangling steps in English (not code)
* [3 points] Step 7: Data wrangling
    * correct use of `filter()`
    * correct use of `group_by()`
    * correct use of `summarise()`
* [2 points] `ggplot()` graph of first names 

# Set Up

```{r, message = F, warning = F}
# load necessary packages
library(tidyverse)
library(mosaic)
library(esquisse)

# Load the BabyNames data into RStudio environment (from dcData package)
data("BabyNames", package = "dcData")
```

# Activity

**Goal**: By the end of the activity, each team will create a graph (like Figure A.1 on p 159) showing ups and downs in the popularity of the names at your table according to the U.S. social security registry.  

For steps 5-7, pick any FIVE names you wish. I choose Arnaldo, Ahmid, Nicho, Jennaveve, and Christy. You may choose these names or any others that you would like. 

### Step 1

*Examine the data you have at hand - for this project the data are the ones available in the `BabyNames` data set. Use at least 2 different functions to examine your data. Then answer the questions: What are the variables are available? What is a case?*

Variables

1. name The given name (character string)
2. sex F or M (character string)
3. count The number of babies given that name and of that sex. (integer)
4. year Year of birth (integer)

Cases

A case is the number of babies of a certain sex given a particular name in a given year.

```{r}
head (x = dcData::BabyNames)
count(x = dcData::BabyNames, sex)
count(x = dcData::BabyNames, year) %>%
  arrange(desc(n)) %>%
    head(n = 10)
select(.data = dcData::BabyNames, name, count) %>%
  arrange(desc(count)) %>%
    head(n = 10)
select(.data = dcData::BabyNames, name, count) %>%
  arrange(desc(count)) %>%
    filter(count == 5)
```

### Step 2

[This should be done by hand on paper, nothing to show in the HTML doc]

### Step 3 

##### What variable(s) from the raw data table do not appear at all in the graph?

The variable `sex` does not appear at all in the graph.

##### What variable(s) in the graph are similar to corresponding variables in the raw data table, but might have been transformed in some way?

The variables `year` and `count` appear as the x- and y-axis, respectively. The variable `name` is filtered down to a select subset of cases, which appear as colored lines within the `year`-`count` space.

### Step 4

##### Have cases been filtered out?

Yes!

##### Have cases been grouped and summarized within groups in any way?

Cases have been grouped by year.

##### Have any new variables been introduced?  If so, what the relationship between the new variables and existing variables?

No new variables have been introduced.

### Step 5

Using the BabyNames data set, filter the data to display just the 5 names that you picked.  Then group the cases by year and name, and count how many total births per year were recorded for each of the names (regardless of sex).  

* Select the subset of rows whose name is one of the five chosen names via `filter(name %in% c(...))`
* Group the rows by year, and within a given year group the rows by name via `group_by(year, name)`
* For each year-name group, sum the count of each row to obtain a total count for that group.

### Step 6

[This should be done by hand on paper, nothing to show in the HTML doc]

### Step 7

```{r}
GlyphReadyForm <- dcData::BabyNames %>%
  filter(name %in% c('Abraham', 'Franklin', 'Woodrow')) %>%
    group_by(year, name) %>%
      summarize(total = sum(count))
GlyphReadyForm
```

### Graph It!

Here's the sample code provided by the book:

```
GlyphReadyForm %>%
  ggplot(aes(x = year, y = total, group = name)) + 
  geom_line(size = 1, alpha = 0.5, aes(color = name)) + 
  ylab("Popularity") + 
  xlab("Year") +
  ggtitle("Popularity of Names Over Time")
```

If you prepared your `GlyphReadyForm` data set correctly, you should just be able to copy and paste this code. Alternatively, you can use `mplot` or `esquisse` to create your graph. If you choose to use these interactive functions remember, do not include the `mplot` or `esquisse` functions in your RMarkdown document. Instead, use these functions in the console and copy the generated code into your RMarkkdown document. In addition, we want lines, not dots. See examples below. 

```{r}
GlyphReadyForm %>%
  ggplot() +
    geom_line(
      alpha     = 0.5,
      linewidth = 1,
      mapping   = aes(x        = year,
                      y        = total,
                      color    = name,
                      linetype = name),
      position  = 'identity',
      stat      = 'identity',
      show.legend=TRUE
    ) +
    xlab('Year') +
    ylab('Popularity') +
    ggtitle('Popularity of Names Over Time')
```

#### Example of Dots versus Lines

We can create a scatter plot for our graph. It's close to what we want, but the line plot that we really want as it is easier to understand.

```
ggplot( data = GlyphReadyForm, aes(x = year, y = total) ) +
  geom_point()+
  aes(colour = name) 
```

If I use the recommended `ggplot()` command provided by the textbook I get something the following.  If you look closely at the two commands, you'll see a number of similarities in the syntax.  It's always a good idea to study code examples that work in order to see what you can use now or in the future.  For example, you might learn from the command in the text how to apply custom labels to your axes!

```
GlyphReadyForm %>%
  ggplot(aes(x = year, y = total, group = name)) + 
  geom_line(size = 1, alpha = 0.5, aes(color = name)) + 
  ylab("Popularity") + 
  xlab("Year")
```